Reproducible research in archaeology using R & rrtools

### 2018-03-07 analysis/paper/paper.Rmd
all slides and text basically done

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)

@@

library(ggrepel)
library(tidyverse)
library(glue)
library(tidytext)
library(scales)
```

@@

## Slide 1: Title slide

@@

## Slide 2: Motivation

@@

## Slide 3: Outline

@@

Secondly, we will review a few case studies of how R is used by archaeologists, to show the potential that using a programming language has for opening and accelerating archaeological research.

@@

##- Slide 4: Current uses of R [signpost slide only]

## Slide 5: Articles in all sciences citing R

```{r all-sciences-citing-r, fig.height=10}

@@

min_y <- 2007
max_y <- 2025
ggplot(cran_cites_all_areas_top_journals_by_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_text_repel(
    data = subset(cran_cites_all_areas_top_journals_by_year, 
                  PY == max(PY)),
    aes(label = str_wrap(SO, 40)),
    size = 4,
    nudge_x = 0,
    hjust = 0,
    segment.color = NA,
    direction = "y"
  ) +
  scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, rep("", length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal") +
  theme_minimal() +
  theme(legend.position="none") +
  ggtitle(str_glue('Percentage of articles per year citing R in top 10 WoS journals ({prettyNum(how_many_articles_in_top_journals_for_cran_cites,",")} articles)'),
          subtitle = "Data from apps.webofknowledge.com,\nusing in SCI-EXPANDED & SSCI for 2007-2017")

@@

First, the areas of science that have seen big growth in the use of R are ecology, evolution and related biological sciences. We also see increases in multi-disciplinary journals such as PeerJ and PLOS One.

Second, the scale of the increases is quite impressive. We see many journals in our sample starting from less than 1% of articles citing R and now have around 20% of articles citing R.

@@

## Slide 6: Archaeology articles citing R

@@

library(ggpmisc)
jas_cites_r <- 
cran_cites_in_archy_journals_per_year %>% 
  filter(SO == "JOURNAL OF ARCHAEOLOGICAL SCIENCE") %>% 
  ggplot(aes(PY,
           prop)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm",
              se = FALSE) +
  stat_poly_eq(aes(label =  paste(..eq.label.., 
                                  ..adj.rr.label..,
                                  sep = "~~~~")),
               formula = y~x, 
               parse = TRUE,
               size = 3,
               label.y.npc = 0.7) +
  stat_fit_glance(label.y.npc = 0.7,
                  size = 3,
                  method = "lm", 
                  method.args = list(formula = y ~ x),
                  geom = "text",
                  aes(label = paste("p-value: ",signif(..p.value.., digits = 4)))) +
  xlab("year") +
  ylab("Proportion of articles citing R") +
  ggtitle(expression(paste("Articles citing R in ", italic("Journal of Archaeological Science")))) +
  theme_bw(base_size = 6)

@@

There are four important details in these plots:

The first detail we see here is that overall, the proportions of articles citing R in archaeology are much smaller than what we see in the ecology journals on the previous slide. The absolute numbers of archaeology articles here are are also very low. Only the _Journal of Archaeological Science_ has more than 10 articles in our sample. We have yet to adopt programming for data analysis in the same way that ecologists have.

The second detail is that we do not see a significant increase over time, unlike the ecology journals that steeply increase over time. Both archaeology and the other sciences we saw on the previous slide show an increase in citing R at around 2013-2014, we do not yet see much of an upward trend in the archaeology journals. Indeed, the first citations of R in several archaeology journals only occured in the last two years. The plot on the upper right of this slide shows a statistical test of change over time for articles in JAS. With a low r-squared value and high p-value for the linear model, we conclude that there is no evidence for an increase in citations over time in this journal.

The fourth detail we see is the distinctive types of journals that have articles citing R. There is a clear focus on scientific archaeology, with _Journal of Archaeological Science_ having the highest total number of articles. We also see geoarchaeology, lithics, and environmental archaeology represented in the journal names. We can investigate this further with statistical analysis of the words in the titles of all the articles in our sample.

## Slide 7: What are archaeologists doing with R?

@@

# subtract ones citing R from ones that do not
library(Hmisc)
wc_archaeology_2007_to_2017_no_R_cites <- 
  wc_archaeology_2007_to_2017 %>% 
  filter(TI %nin% cran_cites_in_archy_journals$TI)

@@

## Slide 8: The unrealised potential of R for reproducible research [signpost slide only]

## Slide 9: The unrealised potential of R for reproducible research [signpost slide only]

@@

## Slide 10: An increasing number of archaeologists are publishing reproducible research

@@

Articles that have these three characteristics are part of a new paradigm in scientific communication, and with these `r number_of_reproducible_articles` articles, we believe we have identified the beginning of this paradigm in archaeology. We think it will not be hard for archaeologists that already use R to make the transition to producing reproducible research by sharing their code and data.

## Slide 11: Our new package, rrtools, and how it helps

@@

- use_compendium:  This function creates a new directory for the research project, and creates in R package within this directory. We use the R package structure as the standard for our directory because it is widely recognised by R users. This means that people know where to look for code, where to look for data, and where to look for dependancies and other information.     
- usethis::use_mit_license: This function creates a copy of the MIT software lisence to the compendium. This is important for two reasons. First, because it tells the reader that the author is happy for their code to be reused by other people. And second, it tells the reader that the author does not take responsibility for any problems that the reader might have when they use the code. This is important for setting the expectations about the relationship between the author and the reader, regarding the use of code.     
- devtools::use_github:  This will initiate the use of the Git version control system. This is the current state-of-the-art for version control of any kind of plain text file, including text, code, data, and images. This will also create a repository on the GitHub website. Using version control is important because it allows the author to time-travel back to earlier states of their analysis, if they take a dead-end path. It is also important for making collaboration smooth because many people can work on one set of files without losing track of what the most recent version is. Plus it is nice to have a remote back-up of our work in case of an emergency. The GitHub repository can be kept private if desired, until the work is published.     
- use_readme_rmd:  This will create a simple document that is usually the first thing a reader sees when they browse the code files. This document is important because it helps to describe to the reader what to expect in the compendium, and it gives details of how to cite the compendium.     
- use_analysis:  This function creates a set of folders according to the best practices for organising a typical reasearch project. We recognise that this file structure wont be perfect for every project. But we know that it will suit many projects quite well, and it will at least make the user think about how to organise their files logically, and according to best practises. On the next slide we can see the folders that this fucntion creates:

## Slide 12: The best-practice directory structure from rrtools

Here is the file structure the rrtools gives us. We wont discuss it in detail, except to point out two important principles that it enforces:

@@

## Slide 13: Advanced reproducibility with rrtools

The last three functions that rrtools gives us are advanced topics that we only summarise briefly here.  ...

@@

>>>>

```{r, setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  fig.height=5,
  fig.width=5,
  fig.path = "../figures/"
)

@@

library(ggrepel)
library(tidyverse)
library(glue)
library(tidytext)
library(scales)
library(ggpmisc)
library(Hmisc)
```

@@

## Slide: Title slide

@@

## Slide: Motivation

@@

## Slide: Outline

@@

Secondly, we will review how R is used by archaeologists, to show the potential that using a programming language has for opening and accelerating archaeological research.

@@

## Slide: Current uses of R [signpost slide only]

## Slide: Articles in all sciences citing R

```{r all-sciences-citing-r}

@@

min_y <- 2007
max_y <- 2025
ggplot(cran_cites_all_areas_top_journals_by_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_text_repel(
    data = subset(cran_cites_all_areas_top_journals_by_year, 
                  PY == max(PY)),
    aes(label = str_wrap(SO, 40)),
    size = 4,
    nudge_x = 0,
    hjust = 0,
    segment.color = NA,
    direction = "y"
  ) +
  scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, rep("", length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  ggtitle(str_glue('Percentage of articles per year citing R in top 10 WoS journals ({prettyNum(how_many_articles_in_top_journals_for_cran_cites,",")} articles)'),
          subtitle = "Data from apps.webofknowledge.com,\nusing in SCI-EXPANDED & SSCI for 2007-2017")

@@

**First**, the areas of science that have seen big growth in the use of R are ecology, evolution and related biological sciences. We also see increases in multi-disciplinary journals such as PeerJ and PLOS One.

**Second**, the scale of the increases is quite impressive. We see many journals in our sample starting from less than 1% of articles citing R and now have around 20% of articles citing R.

@@

## Slide: Archaeology articles citing R

@@

jas_cites_r <- 
cran_cites_in_archy_journals_per_year %>% 
  filter(SO == "JOURNAL OF ARCHAEOLOGICAL SCIENCE") %>% 
  ggplot(aes(PY,
           prop)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm",
              se = FALSE) +
  stat_poly_eq(aes(label =  paste(..eq.label.., 
                                  ..adj.rr.label..,
                                  sep = "~~~~")),
               formula = y~x, 
               parse = TRUE,
               size = 3,
               label.y.npc = 0.7) +
  stat_fit_glance(label.y.npc = 0.7,
                  size = 3,
                  method = "lm", 
                  method.args = list(formula = y ~ x),
                  geom = "text",
                  aes(label = paste("p-value: ",signif(..p.value.., digits = 4)))) +
  xlab("year") +
  ylab("Proportion of articles citing R") +
  ggtitle(expression(paste("Articles citing R in ", italic("Journal of Archaeological Science")))) +
  theme_bw(base_size = 6)

@@

There are three important details in these plots:

The **first** detail we see here is that overall, the proportions of articles citing R in archaeology are much smaller than what we see in the ecology journals on the previous slide. The absolute numbers of archaeology articles here are are also very low. Only the _Journal of Archaeological Science_ has more than 10 articles in our sample. We have yet to adopt programming for data analysis in the same way that ecologists have.

The **second** detail is that we do not see a significant increase over time, unlike the ecology journals that steeply increase over time. Both archaeology and the other sciences we saw on the previous slide show an increase in citing R at around 2013-2014, we do not yet see much of an upward trend in the archaeology journals. Indeed, the first citations of R in several archaeology journals only occured in the last two years. The plot on the upper right of this slide shows a statistical test of change over time for articles in JAS. With a low r-squared value and high p-value for the linear model, we conclude that there is no evidence for an increase in citations over time in this journal.

The **third** detail we see is the distinctive types of journals that have articles citing R. There is a clear focus on scientific archaeology, with _Journal of Archaeological Science_ having the highest total number of articles. We also see geoarchaeology, lithics, and environmental archaeology represented in the journal names. We can investigate this further with statistical analysis of the words in the titles of all the articles in our sample.

## Slide: What are archaeologists doing with R?

@@

# subtract ones citing R from ones that do not

wc_archaeology_2007_to_2017_no_R_cites <- 
  wc_archaeology_2007_to_2017 %>% 
  filter(TI %nin% cran_cites_in_archy_journals$TI)

@@

## Slide: The unrealised potential of R for reproducible research [signpost slide only]

## Slide 9: The unrealised potential of R for reproducible research

![](../figures/Marwick et al Fig 3 screenshot from the RStudio program.png)

@@

## Slide: Real-world example from Barton et al. 2018

![](../figures/example-barton-et-al.png)

Here is a real-world example R Markdown in use, from an archaeology article published earlier this year.

## Slide: An increasing number of archaeologists are publishing reproducible research

@@

Articles that have these three characteristics are part of a new paradigm in scientific communication, and with these `r number_of_reproducible_articles` articles, we believe we have identified the beginning of this paradigm in archaeology. We think it will not be hard for archaeologists that already use R to make the transition to producing reproducible research by sharing their code and data. If this can be achieved, then we can realise some of the potential that R has for making archaeology more open and reproducible.

## Slide: Our new package, rrtools, and how it helps

![](../figures/rrtools-steps-carbon.png)

@@

- **use_compendium:**  This function creates a new directory for the research project, and creates in R package within this directory. We use the R package structure as the standard for our directory because it is widely recognised by R users. This means that people know where to look for code, where to look for data, and where to look for dependancies and other information.     
- **usethis::use_mit_license:** This function creates a copy of the MIT software lisence to the compendium. This is important for two reasons. First, because it tells the reader that the author is happy for their code to be reused by other people. And second, it tells the reader that the author does not take responsibility for any problems that the reader might have when they use the code. This is important for setting the expectations about the relationship between the author and the reader, regarding the use of code.     
- **devtools::use_github:**  This will initiate the use of the Git version control system. This is the current state-of-the-art for version control of any kind of plain text file, including text, code, data, and images. This will also create a repository on the GitHub website. Using version control is important because it allows the author to time-travel back to earlier states of their analysis, if they take a dead-end path. It is also important for making collaboration smooth because many people can work on one set of files without losing track of what the most recent version is. Plus it is nice to have a remote back-up of our work in case of an emergency. The GitHub repository can be kept private if desired, until the work is published.     
- **use_readme_rmd:**  This will create a simple document that is usually the first thing a reader sees when they browse the code files. This document is important because it helps to describe to the reader what to expect in the compendium, and it gives details of how to cite the compendium.     
- **use_analysis:**  This function creates a set of folders according to the best practices for organising a typical reasearch project. We recognise that this file structure wont be perfect for every project. But we know that it will suit many projects quite well, and it will at least make the user think about how to organise their files logically, and according to best practises. On the next slide we can see the folders that this function creates:

## Slide: The best-practice directory structure from rrtools

![](../figures/rrtools-analysis-dir.png)

Here is the file structure the rrtools gives us. There isn't time here to discuss it in detail, except to point out two important principles that it enforces:

@@

## Slide: Advanced reproducibility with rrtools

![](../figures/rrtools-steps-carbon.png)

The last three functions that rrtools gives us are advanced topics that we summarise only very briefly here.

- **use_dockerfile:** This function generates a configuration file that allows another user to recreate the computational environment of the original author. This is important for complex projects that have many pieces of software that might be difficult to install correctly. Sometimes when a piece of software is updates, its behaviour can change, which can alter the results of an analysis. This function helps to protect again this problem.    
- **use_travis:** This function will create a configuration file so that our compendium can be automatically checked each time we update it. This is especially useful when working with collaborators because it helps to identify problems in the analysis before they become too big.    
- **use_testthat:** This creates a template for adding tests for any custom functions used in the analysis. This is important so we can ensure that the functions perform as expected while we work on the anlysis.

These three functions are very widely used in software engineering to give quality control to a package. With our rrtools package we have brought these tools into a research context where they can help researchers to make their analyses easy to work on and easy for others reproduce.

## Slide: Summary

To summarise:

- We have shown a bibliometric analysis that reveals substantial change in the use of the R programming languages in the sciences generally. We saw that ecological and evolutionary sciences have really embraced R in their publications.

- We have also shown that archaeologists have more recently taken up R, but to a much lesser degree, so far, than what we see in the other fields. We've seen that the use of R is most common in more scientific aspects of archaeology.

- We also identified a small sub-set of archaeology papers that use R to make their work fully reproducible. There is a lot of potential for other archaeologists to make their work reproducible, and we think this would be a good thing for archaeology generally.

- We have presented our freely available R package, rrtools, which is designed to make it easy for researchers in archaeology and other fields to make their research reproducible. We hope this will help to realise some of the potential to make archaeological research more open and reproducible.

Thank you.

@@

The rrtools project was developed during the 2017 Summer School on Reproducible Research in Landscape Archaeology at the Freie Universität Berlin (17-21 July), funded and jointly organized by Exc264 Topoi, CRC1266, and ISAAKiel.

Thanks to Daniel Knitter, Patrick Kennedy, Nils Müller-Scheeßel, Martin Hinz, Clemens Schmid, Ricarda Braun, Wojciech Francuzik for their contributions to the package.
